Territoriality and the organization of technology during the Last Glacial Maximum in southwestern Europe

### 2019-09-18 README.Rmd
updated description

[![Last-changedate](https://img.shields.io/badge/last%20change-`r gsub('-', '--', Sys.Date())`-brightgreen.svg)](https://github.com/jmcascalheira/ScaledPiecesVB/commits/master)
[![minimal R version](https://img.shields.io/badge/R%3E%3D-3.2.4-brightgreen.svg)](https://cran.r-project.org/) 
[![Build Status](https://travis-ci.org/jmcascalheira/ScaledPiecesVB.svg?branch=master)](https://travis-ci.org/jmcascalheira/ScaledPiecesVB)
[![Licence](https://img.shields.io/github/license/mashape/apistatus.svg)](http://choosealicense.com/licenses/mit/)
[![ORCiD](https://img.shields.io/badge/ORCiD-/0000--0003--0321--8892-green.svg)](http://orcid.org/0000-0003-0321-8892)

@@

> Cascalheira, J. (2019). _`r Title`_. Name of journal/book, Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>

@@

> Cascalheira, J. (`r format(Sys.Date(), "%Y")`). _Compendium of R code and data for `r Title`_. Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>

@@

You can download the compendium as a zip from from this URL: </archive/master.zip>

Or you can install this compendium as an R package, LGMTechnology, from GitHub with:

@@

>>>>

[![Last-changedate](https://img.shields.io/badge/last%20change-`r gsub('-', '--', Sys.Date())`-brightgreen.svg)](https://github.com/jmcascalheira/ScaledPiecesVB/commits/master)
[![minimal R version](https://img.shields.io/badge/R%3E%3D-3.2.4-brightgreen.svg)](https://cran.r-project.org/) 
[![Licence](https://img.shields.io/github/license/mashape/apistatus.svg)](http://choosealicense.com/licenses/mit/)
[![ORCiD](https://img.shields.io/badge/ORCiD-/0000--0003--0321--8892-green.svg)](http://orcid.org/0000-0003-0321-8892)

@@

> Cascalheira, J. (2019). _`r Title`_. Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/10.31235/osf.io/fgxtu>

@@

> Cascalheira, J. (`r format(Sys.Date(), "%Y")`). _Compendium of R code and data for `r Title`_. Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/10.17605/OSF.IO/YD2VE>

@@

You can download the compendium as a zip from from this URL: <https://github.com/jmcascalheira/LGMtechnology/archive/master.zip>

This repository is organized as an R package using rrtools by Ben Marwick, which can be installed from github with:

```{r gh-installation, eval = FALSE}
# install.packages("devtools")
devtools::install_github("benmarwick/rrtools")
```

There are no actual R functions in this package - all the R code is in the Rmd file. We simply used the R package structure to help manage dependencies, to take advantage of continuous integration for automated code testing, and so I didn't have to think too much about how to organise the files.

To download the package source as you see it on GitHub, for offline browsing, use this line at the shell prompt (assuming you have Git installed on your computer):

```{r eval=FALSE}
git clone https://github.com/jmcascalheira/LGMtechnology.git
```

Once the download is complete, open the `lgmtechnolgoy.Rproj` in RStudio to begin working with the package and compendium files.

The package has a number of dependencies on other R packages, and programs outside of R. Installing these can be time-consuming and complicated, so we've included a packrat directory, which contains the source code for all the packages we depend on. If all works well, these will be installed on your computer when you open `lgmtechnology.Rproj` in RStudio.

@@

### Dependencies:

See the colophon section of the docx file in `analysis/paper` for a full list of the packages that this project depends on.

### Contact:

João Cascalheira, Post-Doc Researcher, ICArEHB
University of Algarve, Campus de Gambelas
8005-139 Faro PORTUGAL
e. jmcascalheira@ualg.pt  w. http://www.icarehb.com/cascalheira

### 2019-09-18 analysis/paper/SOM.Rmd
updated description

---
title: "SOM: Territoriality and the organization of technology during the Last Glacial Maximum in southwestern Europe"
author:
  - João Cascalheira (ICArEHB, University of Algarve, Faro, Portugal) jmcascalheira@ualg.pt
date: '`r Sys.Date()`'
output: 
    bookdown::pdf_document2:
      toc: False
      number_sections: False

@@

## S1 - Blank attribute frequency

>>>>

---
title: "SOM: Territoriality and the organization of technology during the Last Glacial Maximum in southwestern Europe"
author:
  - João Cascalheira (ICArEHB, University of Algarve, Faro, Portugal) jmcascalheira@ualg.pt
output: 
    bookdown::pdf_document2:
      toc: False
      number_sections: False

@@

## S1 - Blanks attribute frequency

### 2019-09-18 analysis/paper/flakes_MCA.Rmd
updated description

cores <- read.csv("../data/raw_data/cores.csv")

cores_tidy <- cores %>%
  drop_na(Length, Width, Thickness, Weight) %>%
  mutate(Elongation = Length / Width) %>%
  mutate(Flattening = Width / Thickness) %>%
  mutate(CoreSection = fct_lump(CoreSection, prop = 0.05)) %>%
  mutate(PlatformType = fct_lump(PlatformType, prop = 0.05)) %>%
  mutate(CortexLoc = fct_lump(CortexLoc, prop = 0.05)) %>%
  mutate(CoreType = fct_lump(CoreType, prop = 0.05)) %>%
  mutate(BlankType = fct_lump(BlankType, prop = 0.05)) %>%
  mutate(Platforms = fct_lump(Platforms, prop = 0.05)) %>%
  mutate(CortexPerc = dplyr::recode(CortexPerc, "75-95%" = "76-100%", "25-75%" ="26-75%", "100%" = "76-100%", "<25%" = "1-25%", ">95%" = "76-100%", "0%" = "0%")) %>%
  mutate(ScarNumber = as.factor(ScarNumber)) %>%
  mutate(ScarNumber = dplyr::recode(ScarNumber, "4" = "4 or more", "5" = "4 or more", "6" = "4 or more", "7" = "4 or more", "8" = "4 or more", "9" = "4 or more"))

# Rename Context column
colnames(cores_tidy)[1] <- "Context"

# Bin continous variables
cores_tidy$ElongFact <- binVariable(cores_tidy$Elongation, bins = 3, method = "natural", labels = c("low", "medium", "high"))
cores_tidy$FlattFact <- binVariable(cores_tidy$Flattening, bins = 3, method = "natural", labels = c("low", "medium", "high"))
cores_tidy$WeightFact <- binVariable(cores_tidy$Weight, bins = 3, method = "natural", labels = c("low", "medium", "high"))
```

@@

cores_table <- cores_tidy %>%
  droplevels()

# Set variables
var_list <- c("CoreType", "Platforms", "CoreSection", "BlankType", "CortexPerc", "Weight",
              "Elongation", "Flattening")

# Get cross-tables by chronology using cross_tb() function
cores_table <- tabmulti(cores_table, "Context", var_list,
                 p.include = FALSE,
                 n.headings = FALSE,
                 means.text.label = "none",
                 freq.text.label = "none")

cores_table <- cores_table %>% 
  as_tibble() %>% 
  select(" " = Variable, `AMB II`, `AMB IV`, `AMB VI`, `PAP 4'00-4'75`, `PAP 4'75-5'25`, `PAP 5'25-6'25`, `VALM`, `VB A`, `VB B`, `VB C`, Total = Overall)

kable(cores_table)

@@

```{r MCA_cores, echo=FALSE}

# Filter specific categories for MCA analysis
cores <- cores_tidy %>%
  select(Context, CortexPerc, CoreType, Platforms, ElongFact, FlattFact, BlankType, PlatformType) %>%
  na.omit() %>%
  droplevels()

@@

mca_res_cores <- MCA(cores, graph = FALSE, quali.sup = 1, method = "Burt")

fviz_screeplot(mca_res_cores, addlabels = TRUE, ylim = c(0, 45))

fviz_mca_var(mca_res_cores, col.var = "contrib",
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             repel = TRUE, invisible = "quali.sup",
             ggtheme = theme_gray())

@@

var_description <- data.frame(
  Dimension = c("Dimension 1", "Dimension 2"),
  Variability = c("23%", "13%"),
  Positive = c("Simple cores with 2 isolated or multiple platforms. Low elongation and high flattening. Extraction of flakes with cortical platforms",
    "Other types of cores with multiple platforms for mixed products"),
  Negative = c("Prismatic cores with 2 opposed platforms. High elongation and low flattening, for elongated products",
  "Simple cores with 1 platform and high % of cortex. Extraction of blades with cortical platforms"
))

kable(var_description)

@@

```{r ANOVA_cores, echo=FALSE}

@@

mca <- as.data.frame(mca_res_cores$ind)
context <- as.data.frame(cores_tidy$Context)

@@

mca <- mca %>%
  mutate(context = `cores_tidy$Context`, "dim1" = coord.Dim.1, "dim2" = coord.Dim.2) %>%
  select(context, dim1, dim2)

@@

# I want to write the letter over each box. Over is how high I want to write it.
over <- 0.1*max(DIM1_boxplot$stats[nrow(DIM1_boxplot$stats),] )

@@

>>>>

blanks <- read.csv("../data/raw_data/blanks.csv")

flake_tidy <- blanks %>%
  filter(Class == "Flake") %>%
  drop_na(Length, Width, Thickness) %>%
  mutate(Elongation = Length / Width) %>%
  mutate(Flattening = Width / Thickness) %>%
  mutate(Profile = fct_lump(Profile, prop = 0.05)) %>%
  mutate(EdgeShape = fct_lump(EdgeShape, prop = 0.05)) %>%
  mutate(CrossSection = fct_lump(CrossSection, prop = 0.05)) %>%
  mutate(PlatformType = fct_lump(PlatformType, prop = 0.05)) %>%
  mutate(CortexLoc = fct_lump(CortexLoc, prop = 0.05)) %>%
  mutate(DorsalPattern = fct_lump(DorsalPattern, prop = 0.05)) %>%
  mutate(Termination = fct_lump(Termination, prop = 0.05)) %>%
  mutate(CortexPerc = dplyr::recode(CortexPerc, "75-95%" = "76-100%", "25-75%" ="26-75%", "100%" = "76-100%", "<25%" = "1-25%", ">95%" = "76-100%", "0%" = "0%")) %>%
  mutate(ScarNumber = as.factor(ScarNumber)) %>%
  mutate(DorsalPattern = dplyr::recode(DorsalPattern, "Unidentifiable" = "Other")) %>%
  mutate(ScarNumber = dplyr::recode(ScarNumber, "4" = "4 or more", "5" = "4 or more", "6" = "4 or more", "7" = "4 or more", "8" = "4 or more", "9" = "4 or more"))

# Rename Context column
colnames(flake_tidy)[1] <- "Context"

# Convert ScarNumber to factor
flake_tidy$ScarNumber <- factor(flake_tidy$ScarNumber)

# Bin continous variables
flake_tidy$ElongFact <- binVariable(flake_tidy$Elongation, bins = 3, method = "natural", labels = c("low", "medium", "high"))
flake_tidy$FlattFact <- binVariable(flake_tidy$Flattening, bins = 3, method = "natural", labels = c("low", "medium", "high"))

@@

flake_table <- flake_tidy %>%
  droplevels()

# Set variables
var_list <- c("PlatformType", "CrossSection", "Profile", "DorsalPattern", "EdgeShape", "Elongation", "Flattening")

# Get cross-tables by chronology using cross_tb() function
flake_table <- tabmulti(flake_table, "Context", var_list,
                 p.include = FALSE,
                 n.headings = FALSE,
                 means.text.label = "none",
                 freq.text.label = "none")

flake_table <- flake_table %>% 
  as_tibble() %>% 
  select(" " = Variable, `AMB II`, `AMB IV`, `AMB VI`, `PAP 4'00-4'75`, `PAP 4'75-5'25`, `PAP 5'25-6'25`, `VALM`, `VB A`, `VB B`, `VB C`, Total = Overall)

kable(flake_table)

@@

```{r MCA_core_exploitation_flake, echo=FALSE}

# Filter specific categories for MCA analysis
core_exploitation_flake <- flake_tidy %>%
  select(Context, DorsalPattern, CortexPerc, ScarNumber, ElongFact) %>%
  na.omit() %>%
  droplevels()

@@

mca_res_core_exploitation_flake <- MCA(core_exploitation_flake, graph = FALSE, quali.sup = 1, method = "Burt")

fviz_screeplot(mca_res_core_exploitation_flake, addlabels = TRUE, ylim = c(0, 45))

fviz_mca_var(mca_res_core_exploitation_flake, col.var = "contrib",
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             repel = TRUE, invisible = "quali.sup",
             ggtheme = theme_gray())

@@

var_description_core_exploitation_flake <- data.frame(
  Dimension = c("Dimension 1", "Dimension 2"),
  Variability = c("34.5%", "17.9%"),
  Positive = c("Initial configuration of cores with different reduction strategies, high presence of cortex on dorsal surfaces, few removals, and low elongation of flakes",
    "Late moment of reduction sequence with no cortex on dorsal surfaces and high number of scars, using bidireccional or multidireccional strategies to extract highly elongated blanks"),
  Negative = c("Moment of the reduction sequence with small amounts of cortex, using bidireccional and unidireccional strategies for the extraction of highly elongated flakes",
  "Moment of the reduction sequence with high amounts of cortex, few removals, unidirectional strategies and low elongation of flakes"
))

kable(var_description_core_exploitation_flake)

@@

```{r ANOVA_core_exploitation_elongated, echo=FALSE}

@@

mca <- as.data.frame(mca_res_core_exploitation_flake$ind)
context <- as.data.frame(core_exploitation_flake$Context)

@@

mca <- mca %>%
  mutate(context = `core_exploitation_flake$Context`, "dim1" = coord.Dim.1, "dim2" = coord.Dim.2) %>%
  select(context, dim1, dim2)

@@

# I want to write the letter over each box. Over is how high I want to write it.
over <- 0.1*max( a$stats[nrow(a$stats),] )

@@

var_description_platform_flake <- data.frame(
  Dimension = c("Dimension 1", "Dimension 2"),
  Variability = c("18.4%", "12.7%"),
  Positive = c("Early moment in the reduction sequence, with high amounts of cortex on the core and percurssion platforms",
    "Early moment in the reduction sequence with high amounts of cortex on the core's debitage surface, with crushed and cortical platforms"),
  Negative = c("Late moment in sequence, with no cortex on dorsal surfaces, high elongation indices and crushed and faceted platforms",
  "Moment in sequence with significant amounts of cortical surface on the debitage surface, using faceted and plain platforms"
))

kable(var_description_platform_flake)

### 2019-09-18 analysis/paper/paper.Rmd
updated description

```{r excpectations, echo=FALSE}

@@

All analyses and data processing were accomplished in R (version `r getRversion()`) [@r_core_team_r_2013]. The complete R code used for all the analysis and visualizations contained in this paper is available at XXXXX. To produce those files the procedures described by Marwick et al. [-@marwick_packaging_2017] for the creation of research compendiums to enhance the reproducibility of research were followed. The files provided contain all the raw data used in our analysis as well as a custom R package [@wickham_r_2015] holding the code to produce all tables and figures. To enable maximum re-use, code is released under the MIT license, data as CC-0, and figures as CC-BY (for more information see [@marwick_computational_2016]).

@@

Another relevant, general, trend is the progressive reduction of platform facetting over time at AMB and PAP, even though, with exception of VALM, assemblages present reduced frequencies of platform preparation. Still, although bifacial thinning flakes have been removed from the analysis, the lower percentages of faceted platforms in latest assemblages at AMB and PAP is probably also indicative of a gradual reduction in the importance of bifacial debitage over time [see @andrefsky_jr_lithics_2005; @parry_expedient_1987 for similar patterns], since the recurrent use of small-scale flaking of striking surfaces to support “on-the-edge” marginal percussion Bradley and Sampson, 1986) is typically associated with bifacial knapping strategies [@stout_late_2014]. On the other hand, platform preparation through facetting has been shown to relate with a purposeful alteration of exterior platform angles in other type of reduction schemes, to control the size and shape of the extracted blanks. One cannot exclude thus that, in some of the studied assemblages, platform facetting was used to control blanks' morphometrics. At VALM, the largest frequency of platform facetting preferentially associated with the production of elongated blanks seems to corroborate this interpretation. Thus, the lower frequencies of platform facetting in the remaining assemblages on the one hand, and their significantly high presence at other Solutrean sites in the portuguese Estremadura [e.g., Casal do Cepo - @zilhao_o_1997] on the other hand, seems to indicate the existence of particular technological solutions in central Portugal that might imply lower degrees of technological knowledge sharing with the southernmost regions of Iberia. Although some aspects of the singularity of VALM might originate from a biased collection of the materials during the excavation, already mentioned by Zilhão [-@zilhao_o_1997], or even from possible differences arising from the single-horizon open-air nature of VALM *contra* the palimpsest-derived assemblages of the remaining sites, the technological dissimilarities detected in this study seem to corroborate the isolation of central Portugal as an independent techno-cultural unit.

@@

As in any other geographies and timeframes, technological variation over the LGM in southern Iberia was the result of a complex relationship among patterns of inheritance, interaction, and regional adaptation. Population interaction within larger territories was certainly a key factor of the adaptive response to the impact of climate and landscape modifications and seem to have functioned through sharing behaviors of stylistic/typological concepts, but also of technological solutions. Concept sharing could have occurred across regions in different acculturation environments, depending on the geographical distance between communities and preferential cultural ties, but also on the impact that local resources could have had in the technological decisions. After all, humans, as other biota, interact with the environment at distinct scales and create self-reinforcing adaptive patterns. The multiple but distinct scales of self-organization during the LGM in southern Iberia, and the distribution of behaviors within and across scales might have been one of the most important generators of resilience for those communities [@cascalheira_cross-scale_2017].

>>>>

```{r expectations, echo=FALSE}

@@

All analyses and data processing were accomplished in R (version `r getRversion()`) [@r_core_team_r_2013]. The complete R code used for all the analysis and visualizations contained in this paper is available at http://www.doi.org/10.17605/OSF.IO/YD2VE. To produce those files the procedures described by Marwick et al. [-@marwick_packaging_2017] for the creation of research compendiums to enhance the reproducibility of research were followed. The files provided contain all the raw data used in our analysis as well as a custom R package [@wickham_r_2015] holding the code to produce all tables and figures. To enable maximum re-use, code is released under the MIT license, data as CC-0, and figures as CC-BY (for more information see [@marwick_computational_2016]).

@@

Another relevant, general, trend is the progressive reduction of platform facetting over time at AMB and PAP, even though, with exception of VALM, assemblages present reduced frequencies of platform preparation. Still, although bifacial thinning flakes have been removed from the analysis, the lower percentages of faceted platforms in latest assemblages at AMB and PAP is probably also indicative of a gradual reduction in the importance of bifacial debitage over time [see @andrefsky_jr_lithics_2005; @parry_expedient_1987 for similar patterns], since the recurrent use of small-scale flaking of striking surfaces to support “on-the-edge” marginal percussion) is typically associated with bifacial knapping strategies [@stout_late_2014]. On the other hand, platform preparation through facetting has been shown to relate with a purposeful alteration of exterior platform angles in other type of reduction schemes, to control the size and shape of the extracted blanks. One cannot exclude thus that, in some of the studied assemblages, platform facetting was used to control blanks' morphometrics. At VALM, the largest frequency of platform facetting preferentially associated with the production of elongated blanks seems to corroborate this interpretation. Thus, the lower frequencies of platform facetting in the remaining assemblages on the one hand, and their significantly high presence at other Solutrean sites in the portuguese Estremadura [e.g., Casal do Cepo - @zilhao_o_1997] on the other hand, seems to indicate the existence of particular technological solutions in central Portugal that might imply lower degrees of technological knowledge sharing with the southernmost regions of Iberia. Although some aspects of the singularity of VALM might originate from a biased collection of the materials during the excavation, already mentioned by Zilhão [-@zilhao_o_1997], or even from possible differences arising from the single-horizon open-air nature of VALM *contra* the palimpsest-derived assemblages of the remaining sites, the technological dissimilarities detected in this study seem to corroborate the isolation of central Portugal as an independent techno-cultural unit.

@@

As in any other geographies and timeframes, technological variation over the LGM in southern Iberia was the result of a complex relationship among patterns of inheritance, interaction, and regional adaptation. Population interaction across larger territories was certainly a key factor of the adaptive response to the impact of climate and landscape modifications and seem to have functioned through sharing behaviors of stylistic/typological concepts, but also of technological solutions. Concept sharing could have occurred across regions in different acculturation environments, depending on the geographical distance between communities and preferential cultural ties, but also on the impact that local resources could have had in the technological decisions. After all, humans, as other biota, interact with the environment at distinct scales and create self-reinforcing adaptive patterns [@sundstromTransdisciplinaryApplicationCrossscale2014]. The multiple but distinct scales of self-organization during the LGM in southern Iberia, and the distribution of behaviors within and across scales might have been one of the most important generators of resilience for those communities [@cascalheira_cross-scale_2017].
